From Reversible Programs to \\ Univalent Universes and Back

### 2017-06-26 upi.tex
working on intro etc

\begin{abstract}
  We establish a close connection between a reversible programming language
  based on type isomorphisms and a formally presented univalent universe. The
  correspondence relates combinators witnessing type isomorphisms in the
  programming language to paths in the univalent universe; combinator
  optimizations in the programming language to 2-paths in the univalent
  universe. The result suggests a simple computational interpretation of paths
  and of univalence in terms of familiar programming constructs whenever the
  universe in question is computable.
\end{abstract}

@@

@@

>>>>

\begin{abstract}
  We establish a close connection between a reversible programming language
  based on type isomorphisms and a formally presented univalent universe. The
  correspondence relates combinators witnessing type isomorphisms in the
  programming language to paths in the univalent universe and combinator
  optimizations in the programming language to 2-paths in the univalent
  universe. The result suggests a simple computational interpretation of paths
  and of univalence in terms of familiar programming constructs whenever the
  universe in question is computable.
\end{abstract}

@@

The proceedings of the 2012 Symposium on Principles of Programming
Languages~\cite{Field:2012:2103656} included two apparently unrelated papers:
\emph{Information Effects} by James and Sabry and \emph{Canonicity for
  2-dimensional type theory} by Licata and Harper. The first paper, motivated by
the physical nature of
computation~\cite{Landauer:1961,PhysRevA.32.3266,Toffoli:1980,bennett1985fundamental,Frank:1999:REC:930275},
proposed a reversible language $\Pi$ in which every program is a type
isomorphism. The second paper, motivated by the topological nature of
computation~\cite{vv06,hottbook}, proposed a judgmental formulation of
intensional dependent type theory with a twice-iterated identity type. During
the conference, it became apparent, at an intuitive and informal level, that the
two papers had strong similarities. Formalizing the precise connection was far
from obvious, however. In this paper we report on a formal connection between
appropriately formulated reversible languages on one hand and univalent
universes on the other.

@@

The first step was a refinement and extension of the
language $\Pi$ with a layer of program transformations between the programs
witnessing type isomorphisms~\cite{Carette2016}. The second crucial result was
the formal presention

Second,
